---
title: "assignment"
author: "Dea Karam"
date: "2026-04-01"
output: 
  html_document:
    theme: flatly
    highlight: kate
    number_sections: no
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Import required packages
```{r}
suppressPackageStartupMessages({
    library(AnnotationHub)
    library(ensembldb)  
    library(GenomicRanges)
    library(MotifDb)
    library(motifmatchr)
    library(TFBSTools)
    library(universalmotif)
    library(Biostrings) 
})
```

## Download the data

For this analysis, I chose the transcription factor REST, a transcriptional repressor best known for silencing neuronal genes in non-neuronal cells.

The ChIP-seq experiment was performed on the human GM12878 cell line, a lymphoblastoid cell line, to identify genomic regions bound by REST.

```{r}
download.file("https://www.encodeproject.org/files/ENCFF428QEI/@@download/ENCFF428QEI.bed.gz", "hs_rest.bed.gz")
peaks <- rtracklayer::import("hs_rest.bed.gz", format = "NarrowPeak")
seqlevelsStyle(peaks) <- "Ensembl" 
```
## Downloading the genome

```{r}
ah <- AnnotationHub()
ah_query <- AnnotationHub::query(ah, c("Homo sapiens", "GRCh38", "2bit", "dna_sm"))
genome_id <- rev(ah_query$ah_id)[1]
genome <- ah[[genome_id]]
genome_seqs <- import(genome)
```

## Loading motifs 

We query the MotifDb database to find motif sequences for REST.

```{r}
motifs <- query(MotifDb, c("REST"))
names(motifs)
```

I chose one of the above motif sequences to proceed with the analysis.

```{r}
motif <- motifs[["Hsapiens-HOCOMOCOv10-REST_HUMAN.H10MO.A"]]
```

```{r, fig.width= 20, fig.height=5}
view_motifs(motif)
```

We then convert the motif seuqnece into the appropriate format.

```{r}
motif2 <- convert_motifs(motif, class="TFBSTools-PWMatrix")
```

## Analysis

### Question 1

"Of all the peaks, what proportion contains a motif for the factor?"

For that, we first run motifmatchr to find the motif instances that are within the REST peaks.

```{r}
moi_peaks <- motifmatchr::matchMotifs(motif2, subject=peaks, genome= genome_seqs, out="positions")[[1]]
moi_peaks
```


```{r}
print(table(overlapsAny(peaks, moi_peaks)))
cat("The total number of peaks that the genome contains:", length(peaks))
```

```{r}
num_peaks <- length(peaks)
peak_has_motif <- overlapsAny(peaks, moi_peaks)
motif_peak_overlaps <- sum(peak_has_motif)
pct_overlaps <- (motif_peak_overlaps/num_peaks) * 100
cat("Of the", num_peaks, "peaks,", motif_peak_overlaps, "(", round(pct_overlaps, 2), "%) contain a motif.\n")
```
### Question 2

"Of all instances of that motif in the genome (or in one chromosome), what proportion is bound by the factor (i.e.
has a peak)?"

For this question, I only searched for motif instances within chromosome 1.

```{r}
chr_seq <- genome_seqs["1"]
peaks_chr1 <- peaks[seqnames(peaks)=="1"]
moi_chr <- motifmatchr::matchMotifs(motif2, subject=chr_seq, genome= genome_seqs, out="positions")
moi_chr
```

In the following code chunk, I simply convert the resulting IRanges object to GRanges.

```{r}
moi_chr <- moi_chr[[1]][[1]]
moi_chr_strand <- mcols(moi_chr)$strand


moi_chr_gr <- GRanges(
  seqnames = "1",
  ranges = IRanges(start = start(moi_chr), end = end(moi_chr)),
  strand = moi_chr_strand
)

```

```{r}
table(overlapsAny(moi_chr_gr, peaks_chr1, type ="within"))
```

```{r}
num_motifs_chr <- length(moi_chr_gr)
motif_in_peak <- overlapsAny(moi_chr_gr, peaks_chr1, type ="within")
ov <- sum(motif_in_peak)
pct_overlaps_chr <- (ov/num_motifs_chr) * 100
cat("Of the", num_motifs_chr, "motif instances,", ov, "(", round(pct_overlaps_chr, 2), "%) overlap a peak\n")
```